---
title: 'Chapter 1: Data cleaning and QC'
output: html_document
date: "2023-07-11"
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE) #prevents R printing code that generated the plot 

```

**Importing peptide and prefeast data**
```{r}

#load packages
library(openxlsx)
library(data.table)
library(tibble)
library(dplyr)
library(tidyverse)

file_path1 <- "Input files/QuantSheet_Peptide_40-225.xlsx"
worksheet_name1 <- "Main"
post_dem <- read.xlsx(file_path1, sheet = worksheet_name1, rows = c(1:9), cols = c(27:132))
post_prot <- read.xlsx(file_path1, sheet = worksheet_name1, rows = c(10:4278), cols = c(1:8, 27:129))

file_path2 <- "Input files/QuantSheet_PreFeaST_40-225.xlsx"
worksheet_name2 <- "PreFeaST_"
pre_dem <- read.xlsx(file_path2, sheet = worksheet_name2, rows = c(1:9), cols = c(9:26, 28:44, 46:62, 64:80, 82:98, 100:116))
pre_prot <- read.xlsx(file_path2, sheet = worksheet_name2, rows = c(10:20581), cols = c(1:26, 28:44, 46:62, 64:80, 82:98, 100:116))

rosmap <- read.xlsx("Input files/Rush_ROSMAP.xlsx")

```

**Filter pre_prot to only show X1s that appear in post_prot**
```{r}

pre_prot <- pre_prot %>% 
  filter(pre_prot$X1 %in% post_prot$X1) 

```

**Pivot proteomic tables and join demographic information**
```{r}

#post-feast data
post_prot <- pivot_longer(post_prot, cols = TMT01_126:TMT06_134C, names_to = "SampleCode", values_to = "Quant")
post_dem <- post_dem %>%
  pivot_longer(-X1, names_to = "SampleCode", values_to = "Val") %>%
  pivot_wider(names_from = X1, values_from = Val)
post_full_data <- left_join(x=post_prot, y=post_dem, by=c("SampleCode"))

#pre-feast data
pre_prot <- pivot_longer(pre_prot, cols = TMT01_126:TMT06_134C, names_to = "SampleCode", values_to = "Quant")
pre_dem <- pre_dem %>%
  pivot_longer(-X1, names_to = "SampleCode", values_to = "Val") %>%
  pivot_wider(names_from = X1, values_from = Val)
pre_full_data <- left_join(x=pre_prot, y=pre_dem, by=c("SampleCode"))

identical(post_full_data$X1, pre_full_data$X1)

```

**Compare pre and post-feast data, make values NaN in post that are NaN in pre**
```{r}

post_full_data2 <- cbind(post_full_data, pre_full_data$Quant) 
post_full_data2$logicvector <- is.na(post_full_data2$"pre_full_data$Quant")
post_full_data2$Quant[post_full_data2$logicvector] <- NA
full_data <- select(post_full_data2, -c("pre_full_data$Quant", logicvector))

```

**Summary table to include/exclude peptides**
```{r}

#any peptides with more than 20% missing values should be excluded from further analysis. We have 102 participants so any peptide with more than 20 missing values will be excluded

NAcounts <- full_data %>%
  group_by(X1) %>%
  summarise(NAcount = sum(is.na(Quant))) 
peptides_to_keep <- NAcounts[-which(NAcounts$NAcount > 20),] #deletes values in NAcounts that exceed 20, keeps all else

```

**Apply NA counts to main data table to delete peptides where NA>20**
```{r}

nice_table <- full_data[full_data$X1 %in% peptides_to_keep$X1, ] #only keep peptide IDs that are in peptides_to_keep
nice_table_prefeast <- pre_full_data[pre_full_data$X1 %in% peptides_to_keep$X1, ] #only keep peptide IDs that are in peptides_to_keep

```

**Get peptide IDs into nice_table and save it as a table**
```{r}

#make a table with the necessary info to make peptide IDs
nametable <- cbind(FullPeptide=nice_table$X1, Sequence=nice_table$Sequence, GNP=nice_table$GeneNamePreferred)
nametable <- data.frame(nametable)

#make peptide IDs
nametable <- nametable %>%
  mutate(after_pipe = ifelse(str_detect(GNP, "\\|"), str_extract(GNP, "(?<=\\|).*"), GNP)) %>% #if there is a | (pipe), remove anything before it
  mutate(before_pipe = ifelse(str_detect(after_pipe, "\\|"), str_replace(after_pipe, "\\|.*$", ""), after_pipe)) %>% #if there is a |, remove anything after it
  mutate(GeneName = str_replace(before_pipe, "GN_", "")) %>% 
  select(-c(GNP, after_pipe, before_pipe)) %>%
  mutate(AALength = str_length(Sequence)) %>% #counts the number of AAs in a peptide
  mutate(firstfour = str_extract(nametable$Sequence, "^.{4}")) %>% #get the first 4 letters of each sequence
  unite(FullPeptideID, c(GeneName, firstfour, AALength), remove=FALSE) #put together gene name, firstfour and length, split by '_'

#distinction between post-mods by adding _B
nametable <- nametable %>%
  unique() %>%
  group_by(FullPeptideID) %>%
  mutate(rank_col = seq_along(FullPeptide)) %>%
  mutate(FullPeptideID = if_else(rank_col == 2, paste0(FullPeptideID, "_B"), FullPeptideID,)) %>% #add _B for top 6 to show that they are different
  ungroup()

nice_table <- merge(nametable, nice_table, by.x = "FullPeptide", by.y = "X1", all = TRUE) #merge table with new ranks 

write_csv(nice_table, "Output files/nice_table.csv")

```

**Join with extra RADC variables**
````{r}

#get dataframe of unique sampleIDs
sampleIDs <- as.data.frame(nice_table$sampleID_Client)
sampleIDs <- sampleIDs %>% 
  rename(sampleID = "nice_table$sampleID_Client") %>%
  unique()

#select columns wanted from big rosmap spreadsheet
rosmap_small <- rosmap %>%
  select(projid, cogn_global_lv, gpath, pmi, braaksc, educ, cts_mmse30_lv)

#filter for shared peptides and merge to nice_table
rosmap_small$projid <- as.character(rosmap_small$projid)
filtered_rosmap <- semi_join(rosmap_small, sampleIDs, by = c("projid" = "sampleID"))
filtered_rosmap2 <- filtered_rosmap[rep(seq_len(nrow(filtered_rosmap)), each = 3696), ]
nice_table2 <- merge(x = filtered_rosmap, y = nice_table, by.x = "projid", by.y = "sampleID_Client")

write_csv(nice_table2, "Output files/nice_table2.csv")

```

