---
title: "Chapter 2: Linear model"
output: html_document
date: "2023-07-11"
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE) 

```

**Import pre-processed data**
```{r}

nice_table <- read.csv("Output files/nice_table.csv")
nice_table2 <- read.csv("Output files/nice_table2.csv")
rosmap <- read.xlsx("Input files/Rush_ROSMAP.xlsx")

```

**Run class1 discrete linear models**
```{r}

#install.packages("broom")
library(broom)

#change class1 from categorical to factor, age from factor to numeric, and gender from chr to factor
nice_table2$class1 <- factor(nice_table2$class1)
nice_table2$age <- as.numeric(nice_table2$age)
nice_table2$gender <- factor(nice_table2$gender)

#lm1 - Control reference factor model
nice_table2$class1 <- relevel(nice_table2$class1, ref = "Control")

linmodcont <- nice_table2 %>% 
  nest_by(FullPeptide) %>%
  mutate(LinearModel = list(lm(Quant ~ class1 + age + gender + pmi + educ, data = data, na.action = na.omit))) %>%
  summarise(tidy(LinearModel))
linmodcont <- linmodcont %>%
  mutate(Reference = nice_table2$class1[4]) %>%
  mutate(Reference = case_when(str_starts(term, "class1") ~ "Control", term == "genderm" ~ "Female", term == "age" ~ " ", term == "educ" ~ " ", term == "pmi" ~ " ", term == "(Intercept)" ~ " ", TRUE ~ Reference)) %>% #add comparisons
  relocate(Reference, .before=estimate) #move reference column to start to make it clearer

#lm2 - Dementia reference factor
nice_table2$class1 <- relevel(nice_table2$class1, ref = "Dementia")

linmoddem <- nice_table2 %>%
  nest_by(FullPeptide) %>%
  mutate(LinearModel = list(lm(Quant ~ class1 + age + gender + pmi + educ, data = data, na.action = na.omit))) %>%
  summarise(tidy(LinearModel))
linmoddem <- linmoddem %>%
  mutate(Reference = nice_table2$class1[1]) %>%
  mutate(Reference = case_when(str_starts(term, "class1") ~ "Dementia", term == "genderm" ~ "Female", term == "age" ~ " ", term == "educ" ~ " ", term == "pmi" ~ " ", term == "(Intercept)" ~ " ", TRUE ~ Reference)) %>% #add comparisons
  relocate(Reference, .before=estimate) #move reference column to start to make it clearer

#lm3 - Resilient reference factor
nice_table$class2 <- relevel(nice_table2$class1, ref = "Resilient")

linmodres <- nice_table2 %>%
  nest_by(FullPeptide) %>%
  mutate(LinearModel = list(lm(Quant ~ class1 + age + gender + pmi + educ, data = data, na.action = na.omit))) %>%  
  summarise(tidy(LinearModel))
linmodres <- linmodres %>%
  mutate(Reference = nice_table2$class1[2]) %>%
  mutate(Reference = case_when(str_starts(term, "class1") ~ "Resilient", term == "genderm" ~ "Female", term == "age" ~ " ", term == "educ" ~ " ", term == "pmi" ~ " ", term == "(Intercept)" ~ " ", TRUE ~ Reference)) %>% #add comparisons  
  relocate(Reference, .before=estimate) #move reference column to start to make it clearer


```

**Making summary tables for explanatory variables by peptide - class1**
```{r}

#class1 table - make summary tables
class1tablecont <- linmodcont %>%
  filter(str_detect(term, '^class1')) #extracts all class comparisons

class1tabledem <- linmoddem %>%
  filter(str_detect(term, '^class1')) %>%
  filter(!(str_detect(term, '^class1C'))) #extracts dem vs res and dem vs frail
 
class1tableres <- linmodres %>%
  filter(str_detect(term, '^class1F')) #extracts res vs frail only
 
#collate summary tables
class1tableall <- rbind(class1tablecont, class1tabledem, class1tableres)
class1tableall <- class1tableall %>%
  arrange(FullPeptide) #all comparisons of one peptide shown together (6 for each)

class1tableall <- class1tableall %>%
 mutate(padjBH = p.adjust(p.value, method = "BH")) %>%
  mutate(significance = ifelse(padjBH < 0.05, "significant", "ns"))

#filter for most significant p-values
class1tablesignificant <- class1tableall %>%
  filter(significance == "significant") 
class1tableall <- merge(x = class1tableall, y = nice_table[ , c("FullPeptideID", "FullPeptide")], by = "FullPeptide", all = TRUE)
class1tableall <- class1tableall %>%
  unique() %>%
  na.omit()

write_csv(class1tableall, "Output files/class1tableall.csv")

class1tablesignificant <- merge(x = class1tablesignificant, y = nice_table[ , c("FullPeptideID", "FullPeptide")], by = "FullPeptide", all = TRUE)
class1tablesignificant <- class1tablesignificant %>%
  unique() %>%
  na.omit() %>%
  arrange(padjBH)

write_csv(class1tablesignificant, "Output files/class1tablesignificant.csv")

```

**Making summary tables for explanatory variables by peptide - age**
```{r}

agetable <- linmodcont %>%
  filter(term == "age") %>%
  mutate(padjBH = p.adjust(p.value, method = "BH")) %>%
  mutate(significance = ifelse(padjBH < 0.05, "significant", "ns"))

#filter for most significant p-values
agetablesignificant <- agetable %>%
  filter(significance == "significant") %>%
  arrange(padjBH) #sorts in order of smallest to largest p value (most significant comparisons shown first)

agetable <- merge(x = agetable, y = nice_table[ , c("FullPeptideID", "FullPeptide")], by = "FullPeptide", all = TRUE)
agetable <- agetable %>%
  unique() %>%
  na.omit()

write_csv(agetable, "Output files/agetable.csv")

agetablesignificant <- merge(x = agetablesignificant, y = nice_table[ , c("FullPeptideID", "FullPeptide")], by = "FullPeptide", all = TRUE)
agetablesignificant <- agetablesignificant %>%
  unique() %>%
  na.omit() %>%
  arrange(padjBH)
downregulated_peptides_age <- sum(agetablesignificant$estimate < 0)

write_csv(agetablesignificant, "Output files/agetablesignificant.csv")

```

**Making summary tables for explanatory variables by peptide - gender**
```{r}

gendertable <- linmodcont %>%
  filter(term == "genderm") %>%
  mutate(padjBH = p.adjust(p.value, method = "BH")) %>%
  mutate(significance = ifelse(padjBH < 0.05, "significant", "ns"))

#filter for most significant p-values
gendertablesignificant <- gendertable %>%
  filter(significance == "significant") %>%
  arrange(padjBH) #sorts in order of smallest to largest p value (most significant comparisons shown first)

gendertable <- merge(x = gendertable, y = nice_table[ , c("FullPeptideID", "FullPeptide")], by = "FullPeptide", all = TRUE)
gendertable <- gendertable %>%
  unique() %>%
  na.omit()

write_csv(gendertable, "Output files/gendertable.csv")

gendertablesignificant <- merge(x = gendertablesignificant, y = nice_table[ , c("FullPeptideID", "FullPeptide")], by = "FullPeptide", all = TRUE)
gendertablesignificant <- gendertablesignificant %>%
  unique() %>%
  na.omit() %>%
  arrange(padjBH)
downregulated_peptides_gender <- sum(gendertablesignificant$estimate < 0)

write_csv(gendertablesignificant, "Output files/gendertablesignificant.csv")

```

**Making summary tables for explanatory variables by peptide - pmi**
```{r}

pmitable <- linmodcont %>%
  filter(term == "pmi") %>%
  mutate(padjBH = p.adjust(p.value, method = "BH")) %>%
  mutate(significance = ifelse(padjBH < 0.05, "significant", "ns"))

#filter for most significant p-values
pmitablesignificant <- pmitable %>%
  filter(significance == "significant") %>%
  arrange(padjBH) #sorts in order of smallest to largest p value (most significant comparisons shown first)

pmitable <- merge(x = pmitable, y = nice_table[ , c("FullPeptideID", "FullPeptide")], by = "FullPeptide", all = TRUE)
pmitable <- pmitable %>%
  unique() %>%
  na.omit()

write_csv(pmitable, "Output files/pmitable.csv")

pmitablesignificant <- merge(x = pmitablesignificant, y = nice_table[ , c("FullPeptideID", "FullPeptide")], by = "FullPeptide", all = TRUE)
pmitablesignificant <- pmitablesignificant %>%
  unique() %>%
  na.omit() %>%
  arrange(padjBH)
downregulated_peptides_pmi <- sum(pmitablesignificant$estimate < 0)

write_csv(pmitablesignificant, "Output files/pmitablesignificant.csv")

```

**Making summary tables for explanatory variables by peptide - education**
```{r}

eductable <- linmodcont %>%
  filter(term == "educ") %>%
  mutate(padjBH = p.adjust(p.value, method = "BH")) %>%
  mutate(significance = ifelse(padjBH < 0.05, "significant", "ns"))

#filter for most significant p-values
eductablesignificant <- eductable %>%
  filter(significance == "significant") %>%
  arrange(padjBH) #sorts in order of smallest to largest p value (most significant comparisons shown first)

eductable <- merge(x = eductable, y = nice_table[ , c("FullPeptideID", "FullPeptide")], by = "FullPeptide", all = TRUE)
eductable <- eductable %>%
  unique() %>%
  na.omit()

write_csv(eductable, "Output files/eductable.csv")

eductablesignificant <- merge(x = eductablesignificant, y = nice_table[ , c("FullPeptideID", "FullPeptide")], by = "FullPeptide", all = TRUE)
eductablesignificant <- eductablesignificant %>%
  unique() %>%
  na.omit() %>%
  arrange(padjBH)
downregulated_peptides_educ <- sum(eductablesignificant$estimate < 0)

write_csv(eductablesignificant, "Output files/eductablesignificant.csv")

```

**Run continuous lm**
```{r}

#continuous linear model with global pathology and global cognition (+age, sex, pmi and education)

#run lm
cogvspath <- nice_table2 %>% 
  nest_by(FullPeptide) %>%
  mutate(LinearModelCont = list(lm(Quant ~ cogn_global_lv + gpath + age + gender + pmi + educ, data = data, na.action = na.omit))) %>%
  summarise(tidy(LinearModelCont, conf.int = TRUE))

#Makes the reference variables look nice
cogvspath$Reference <- ifelse(cogvspath$term == 'genderm', "Female", " ")
cogvspath <- cogvspath %>% 
  relocate(Reference, .before=estimate) #move reference column to start to make it clearer

#filter for most significant p-value (cognition)
cogvspath_all <- cogvspath %>%
  filter(term == "gpath" | term == "cogn_global_lv") %>% #should separate these
  mutate(padjBH = p.adjust(p.value, method = "BH")) 

cogvspath_all <- merge(x = cogvspath_all, y = nice_table[ , c("FullPeptideID", "FullPeptide")], by = "FullPeptide", all = TRUE)
cogvspath_all <- cogvspath_all %>%
  unique() %>%
  na.omit() %>%
  mutate(significance = ifelse(padjBH < 0.05, "significant", "ns"))

write_csv(cogvspath_all, "Output files/cogvspath_all.csv")

cogvspath_significant <- cogvspath_all %>%
  mutate(significance = ifelse(padjBH < 0.05, "significant", "ns")) %>%
  filter(significance == "significant") %>%
  arrange(padjBH) 
cogvspath_significant <- merge(x = cogvspath_significant, y = nice_table[ , c("FullPeptideID", "FullPeptide")], by = "FullPeptide", all = TRUE)
cogvspath_significant <- cogvspath_significant %>%
  unique() %>%
  na.omit() %>%
  arrange(padjBH)

```

**Making summary tables for explanatory variables by peptide - global cognition**
```{r}

#cognition table - make summary table
cognition_cont <- cogvspath %>%
  filter(term == "cogn_global_lv") %>%
  mutate(padjBH = p.adjust(p.value, method = "BH")) %>%
  mutate(significance = ifelse(padjBH < 0.05, "significant", "ns"))

#filter for most significant p-values
cognition_cont_significant <- cognition_cont %>%
  filter(significance == "significant") %>%
  arrange(padjBH) 

cognition_cont <- merge(x = cognition_cont, y = nice_table[ , c("FullPeptideID", "FullPeptide")], by = "FullPeptide", all = TRUE)
cognition_cont <- cognition_cont %>%
  unique() %>%
  na.omit()

write_csv(cognition_cont, "Output files/cognition_cont.csv")

cognition_cont_significant <- merge(x = cognition_cont_significant, y = nice_table[ , c("FullPeptideID", "FullPeptide")], by = "FullPeptide", all = TRUE)
cognition_cont_significant <- cognition_cont_significant %>%
  unique() %>%
  na.omit() %>%
  arrange(padjBH)
downregulated_peptides_cogcont <- sum(cognition_cont_significant$estimate < 0)
upregulated_peptides_cogcont <- sum(cognition_cont_significant$estimate > 0)

write_csv(cognition_cont_significant, "Output files/continuous_linear_model_significantbycognition.csv")

```

**Making summary tables for explanatory variables by peptide - global pathology**
```{r}

#pathology table - make summary table
pathology_cont <- cogvspath_all %>%
  filter(term == "gpath") %>%
  mutate(padjBH = p.adjust(p.value, method = "BH")) %>%
  mutate(significance = ifelse(padjBH < 0.05, "significant", "ns"))

pathology_cont_significant <- pathology_cont %>%
  filter(significance == "significant") %>%
  arrange(padjBH) 

pathology_cont <- merge(x = pathology_cont, y = nice_table[ , c("FullPeptideID", "FullPeptide")], by = "FullPeptide", all = TRUE)
pathology_cont <- pathology_cont %>%
  unique() %>%
  na.omit()

write_csv(pathology_cont, "Output files/pathology_cont.csv")

```

**Upset plot - discrete lm**
```{r}

#install.packages('UpSetR')
library(UpSetR)

#create plot
pdf(file = "Figures/UpSetPlot_discrete_new.pdf", width = 8, height = 5) 

#create contrasts column
class1tablesignificant <- class1tablesignificant %>% 
  unite(contrasts, c(term, Reference), remove = FALSE)

write_csv(class1tablesignificant, "Output files/class1tablesignificant.csv")

class1tableall <- class1tableall %>% 
  unite(contrasts, c(term, Reference), remove = FALSE)

write_csv(class1tableall, "Output files/class1tableall.csv")

#pivot from long --> wide
upset_plot <- class1tablesignificant %>% 
  select(FullPeptideID, contrasts, padjBH) %>%
  pivot_wider(names_from = contrasts, values_from = padjBH, values_fill = NA)

list_gen <- function(col){
  upset_plot %>%
    filter(!is.na(col)) %>%
    pull(FullPeptideID)
}

sig_list <- apply(upset_plot[,-1], 2, list_gen)

names(sig_list) <- c("Dementia vs Control", "Frail vs Resilient", "Frail vs Dementia", "Resilient vs Dementia", "Resilient vs Control", "Frail vs Control") #change names for plot

upset(fromList(sig_list), order.by = "freq", nsets = 7, nintersects = 40)

dev.off() 

```

**Upset plot - continuous lm**
```{r}

#install.packages('UpSetR')
library(UpSetR)

#create plot
pdf(file = "Figures/UpSetPlot_cont.pdf", width = 8, height = 5) 

#create contrasts column
cogvspath_significant <- cogvspath_significant %>% 
  unite(contrasts, c(term, Reference), remove = FALSE)

cogvspath_all <- cogvspath_all %>% 
  unite(contrasts, c(term, Reference), remove = FALSE)

#pivot from long --> wide
upset_plot_cont <- cogvspath_significant %>% 
  select(FullPeptideID, contrasts, padjBH) %>%
  pivot_wider(names_from = contrasts, values_from = padjBH, values_fill = NA)

list_gen_cont <- function(col){
  upset_plot_cont %>%
    filter(!is.na(col)) %>%
    pull(FullPeptideID)
}

sig_list_cont <- apply(upset_plot_cont[,-1], 2, list_gen_cont)

names(sig_list_cont) <- c("Global pathology Z score", "Global cognition Z score") #change names for plot

upset(fromList(sig_list_cont), order.by = "freq", nsets = 7, nintersects = 40)

dev.off()

```
